Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-07 numerical.tex

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_modes_iso_kx10_analytic.ps}
  \caption{Growth rate, $\sigma$, vs.\ osciallation frequency, $\omega$, for
  VSI modes with wavenumber $k_x = \khat/H=10/H$ in the fiducial disk model
    with rapid cooling, $\beta=10^{-3}$. %  a disk with 
    % $(\gamma,\Gamma)=(1.4,1.011)$, $(p,q, h)=(-1.5,-1,0.1)$ and
    % $\beta=10^{-3}$. 
We investigate the effect of vertical box size, from $\zmax=3H$ (\emph{blue triangles}) to $7H$ (\emph{green x's})
    with a free surface boundary ($\Delta P=0$).  For $\zmax=5H$  (black dots) we also compare free surface (\emph{black dots}) and rigid (\emph{red plusses}), finding little difference.
      The fundamental mode is marked by `f'. Lines show the analytic predictions for
      isothermal perturbations ($\beta = 0$, \emph{dashed}) and for $\beta = 10^{-3}$ (\emph{dotted}).
      See text for discussion.
   % are computed from the analytic dispersion relation 
    %with thermal relaxation Eq. \ref{relax_disp} (dotted), and that for
    %isothermal perturbations, Eq. \ref{simple_growth} (dashed). 
    \label{compare_modes_iso_kx10} 
  }
\end{figure}

@@

\subsubsection{Case study: $\khat=10$}\label{k10}
Fig. \ref{compare_modes_iso_kx10} maps eigenvalues -- growth rates and
wave frequencies -- of unstable 
modes for $\khat=10$.  Each symbol 
corresponds to a different order mode, i.e.\ a different number of vertical node crossings.  Lower order modes
have lower (in magnitude) frequencies for a given radial wavenumber, as expected for inertial-gravity waves (see \S\ref{stable_novshear}).

 We compare our numerical results to the analytic dispersion relations for 
$\beta = 0$ and $10^{-3}$, from Eqs.\ \ref{simple_growth} and \ref{relax_disp}, respectively.   The analytic models also have discrete modes, which lie on the continuous curves that are plotted for clarity in Fig. \ref{compare_modes_iso_kx10}.   While the analytic models have many simplifications, they crucially lack an artificial vertical surface (as they include an infinite vertical domain).

For our fiducial case (black dots), the lower order and lower frequency modes closely follow the analytic prediction with growth rates increasing with $|\omega|$, the magnitude of the frequency.  However, for $|\omega| \gtrsim 4 h \OmK$ the growth rate declines with $|\omega|$.  This break from the analytic prediction is not due the choice of boundary condition, as we demonstrate by considering rigid boundaries (red plusses).

Rather, the decline in growth rate for high order models is due to the vertical box size, as seen by comparing the $\zmax/H = 3, 5$ and 7 cases in Fig. \ref{compare_modes_iso_kx10}.  Larger boxes give better agreement with the analytic theory, which has infinite vertical extent.  Larger boxes include regions of larger vertical shear (see Eqs. \ref{vshear_thin} and \ref{vertical_shear_ex}), which higher order modes can tap to give larger growth rates.   For the $\zmax = 7 H$ case, we begin to see another branch of modes with the highest growth rates at $\omega \sim -6 h\OmK$. This branch contains the  ``surface modes" described below.

@@

Fig. \ref{lowfreq_eigenfunc} shows 
eigenfunctions $W = \delta P/\rho$ and $\delta   v_z$ for the lowest order fundamental mode.  
Fig. \ref{lowfreq_eigenfunc_2d} maps the pressure perturbation and meridional flow,
scaled by $\sqrt{\rho}$ to reflect the contribution to kinetic energy. Due to the stretched $x$ axis, 
radial velocities are in fact typically much 
smaller than vertical velocities, as expected for a vertically elongated, incompressible mode. 
Most of the kinetic energy is contained within $\sim 2H$ of the midplane due to the density 
stratification.

>>>>

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_modes_iso_kx10_analytic.ps}
  \caption{Growth rate, $\sigma$, vs.\ osciallation frequency, $\omega$, for
  VSI modes with wavenumber $k_x = \khat/H=10/H$ in the fiducial disk model
    with rapid cooling, $\beta=10^{-3}$. %  a disk with 
    % $(\gamma,\Gamma)=(1.4,1.011)$, $(p,q, h)=(-1.5,-1,0.1)$ and
    % $\beta=10^{-3}$. 
We investigate the effect of vertical box size, from $\zmax=3H$
(\emph{blue triangles}) to $7H$ (\emph{green} x's)
    with a free surface boundary ($\Delta P=0$).  For $\zmax=5H$  (black dots) we also compare free surface (\emph{black dots}) and rigid (\emph{red plusses}), finding little difference.
      The fundamental mode is marked by `f'. Lines show the analytic predictions for
      isothermal perturbations ($\beta = 0$, \emph{dashed}) and for $\beta = 10^{-3}$ (\emph{dotted}).
      See text for discussion.
   % are computed from the analytic dispersion relation 
    %with thermal relaxation Eq. \ref{relax_disp} (dotted), and that for
    %isothermal perturbations, Eq. \ref{simple_growth} (dashed). 
    \label{compare_modes_iso_kx10} 
  }
\end{figure}

@@

\subsubsection{Case study: $\khat=10$}\label{k10}
Fig. \ref{compare_modes_iso_kx10} maps the growth rates $\sigma$ and
wave frequencies $\omega$ of unstable 
modes for $\khat=10$.  Each symbol 
corresponds to a different order mode, i.e.\ a different number of vertical node crossings.  Lower order modes
have smaller $|\omega|$ for fixed radial wavenumber, as expected for
inertial-gravity waves (see \S\ref{stable_novshear}).

We compare our numerical results to the analytic dispersion relations for 
$\beta = 0$ and $10^{-3}$, from Eqs.\ \ref{simple_growth} and
\ref{relax_disp}, respectively.   The analytic models also have
discrete modes, which lie on the continuous curves that are plotted
for clarity in Fig. \ref{compare_modes_iso_kx10}.   While the analytic
models have many simplifications, they crucially lack an artificial
vertical surface (as they assume an infinite vertical domain).

For our fiducial case (black dots), the lower order and lower
frequency modes closely follow the analytic prediction with growth
rates increasing with $|\omega|$. %, the magnitude of the frequency.
However, for $|\omega| \gtrsim 4 h \OmK$ the growth rate declines with
$|\omega|$.  This break from the analytic prediction is not due the
choice of boundary condition, as we demonstrate by considering rigid
boundaries (red plusses).

Rather, the decline in growth rate for high order models is due to the
vertical box size, as seen by comparing the $\zmax/H = 3, 5$ and 7
cases in Fig. \ref{compare_modes_iso_kx10}. Increasing $\zmax$ give
better agreement with the analytic theory, which have $\zmax \to  
\infty$. Larger boxes include regions of larger vertical shear (see
Eqs. \ref{vshear_thin} and \ref{vertical_shear_ex}), which higher
order modes can tap to give larger growth rates. For the $\zmax = 7
H$ case, we begin to see another branch of modes with the highest
growth rates at $\omega \sim -6 h\OmK$. This branch contains the
`surface modes' described below.

@@

Fig. \ref{lowfreq_eigenfunc} shows 
eigenfunctions $W = \delta P/\rho$ and $\delta   v_z$ for the lowest order fundamental mode.  
Fig. \ref{lowfreq_eigenfunc_2d} maps the pressure perturbation and meridional flow,
scaled by $\sqrt{\rho}$ to reflect the contribution to kinetic
energy. Notice the stretched $x$ axis. Radial velocities are in fact typically much 
smaller than vertical velocities, as expected for a vertically
elongated, anelastic mode \citepalias{nelson13}. Most of the kinetic
energy is contained within $\sim 2H$ of the midplane due to the
density stratification.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorW_iso} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  1cm]{figures/eigenvectorvz_iso}
  \caption{Pseudo-enthalpy perturbation $W=\delta P/\rho$ (top) and
    vertical velocity perturbation $\delta v_z$ of the fundamental VSI with 
    $\khat=10$ in the fiducial disk with rapid thermal relaxation
    $\beta=10^{-3}$. The real (imaginary) parts of $W$ and $\delta
    v_z$ are plotted as solid (dotted) lines.  
    \label{lowfreq_eigenfunc}
\end{figure}
